Contracts for unions and intersections are hard for Nickel

### 2021-04-13 paper.tex
Add proper intro, small fixes

\title{Union and intersection contracts: theory vs practice}
\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}

@@

\begin{abstract}
 In this paper we do some cool stuff.
\end{abstract}

@@

The explorative work presented in this paper is motived by the development of
the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate them. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages\cite{repology}.  Nickel targets build systems
and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
there is no hesitation in incorporating recent research or doing it ourselves,
the design decisions are always guided by the practical industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, such as the
infrastructure as code paradigm prevailing today in cloud deployment. When the
complexity of a codebase grows, static types becomes attractive again.  For
reusable code — that is, library functions —, static types are specifically
adapated, and bring all the usual benefits of early error detection, code
robustness, better code structure, documentation, and so on.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
which mixes both static and dynamic typing. Unlike gradually-typed languages
like TypeScript, we do not seek to build a complex type system that tries hard
to accept most of the idioms naturally arising in dynamically typed code.
Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing functions operating on generic data easy,
but may require using untyped code for more exotic expressions.  In any case,
this is not a surrender, as Nickel provides a complementary mechanism for more
advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-caluclus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \textit{union types}.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data types [references],
also called \textbf{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In dynamically typed languages, it is customary to directly store different kind
of data in one variable without further ceremony. The runtime system needs to
attach type tags to values anyway, which can usually be observed using functions
like JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
unions} has already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, a lot of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But as a configuration language, Nickel
does not evolve in the abstract. It aims at the best interoperability possible
with JSON. It is designed to generate data for other softwares to consume, and
tries as much as possible to be expressive enough to encode data schemas at the
language level thanks to contracts. If these schemas use untagged union, then
this is a strong incentive to support them at the type level, or at least, at
the contract level.

@@

\begin{lstlisting}
type sum = Int of int | Str of string
\end{lstlisting}

\begin{lstlisting}
{ tag : <Int>, val : Num} | { tag : <Str>, val : Str}
\end{lstlisting}

@@

\begin{lstlisting}
switch x.tag {
  A => x.val + 1,
  B => then strings.fromNum x.val
}
\end{lstlisting}

Flow typing, implemented in Racket\cite{FlowTypingRacket1, FlowTypingRacket2},
Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
TypeScript\cite{FlowTypingTypeScript}, solves the issue by being capable of
refining appropriately the type information in each branch. One can imagine
doing the same in Nickel. Thus equipped with untagged unions and a typechecker
smart enough, classic tagged unions can be simulated, but with the additional
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of union idioms
appearing in the wild.  Thus, adding union types to Nickel sounds appealing.

>>>>

\title{Contracts for unions and intersections are hard for Nickel}
\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}

@@

\begin{abstract}
This paper reports an attempt to incorporate union and intersection types in
Nickel, a configuration programming language with a gradual type system and
first-class contracts. While the end result looks appealing, it turns out the
concrete impact on both the design and the implementation of the language is
unexpectedly high. We review the issues raised in particular by the
implementation of contracts for unions and intersections, and why we think the
cost ends up too high for the benefit. We conclude by suggesting different leads
to represent and type unions.
\end{abstract}

@@

Build systems, package managers, operating systems, cloud infrastructures, and
web services are examples of modern complex software systems that require an
extensive and non-trivial configuration in order to make them adapted to each
different use-case. To manage the growing complexity that is then offloaded to
configurations, the declarative approach has become more and more popular,
illustrated for example by the infrastructure-as-code paradigm. This leads
configuration to play an important role in critical aspects of software
engineering, including security, availability, and maintainability.

However, static text-based configuration alone is falling short of expressivity,
and is seldom sufficient. Data cannot be transformed, combined nor shared,
resulting at best in boilerplate and duplication of information, or at worst, in
data being invalid or inconsistent. Data validation isn’t supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
done. Correctly configuring a modern system is hard and failures may have
substantial negative consequences.

This is why the authors have been developping the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while still providing features for writing correct
code and generating valid data. The focus is first and foremost practical:
Nickel originated as a rethink of the language of the Nix package manager, used
to describe one of the biggest sets of software packages\cite{repology}.  Nickel
targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well. While we are keen on incorporating existing or doing new
research when it solves a well-defined problem, the general design goals are
always guided by the practical industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, in particular with
the infrastructure as code paradigm prevailing today in cloud deployment. When
the complexity of a codebase grows, static types becomes attractive again.  For
reusable code — that is, library functions —, static types are specifically
adapated, and bring all the usual benefits of early error detection, code
robustness, better code structure, documentation, and so on.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
which mixes both static and dynamic typing. Unlike gradually-typed languages
like TypeScript, we do not seek to build a complex type system that tries hard
to accept most of the idioms naturally arising in dynamically typed code.
Rather, we choose to provide a reasonably expressive type system with good
inference properties, which makes typing functions operating on generic data easy,
but may require using untyped code for more exotic expressions.  In any case,
this is not a surrender, as Nickel provides a complementary mechanism for more
advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-caluclus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \emph{union types}.

@@

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data types [references],
also called \emph{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In dynamically typed languages, it is customary to directly store different kind
of data in one variable without further ceremony. The runtime system needs to
attach type tags to values anyway, which can usually be observed using functions
like JavaScript's \verb+typeof+. Using so-called \emph{untagged unions} has
already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, a lot of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. However, Nickel is a configuration
language, whose contracts and types system is intended to be expressive enough
to encode data schemas encoutered in practice. Since these schemas do use
untagged unions, there is a strong incentive to support untagged uniohns at the
type level, or at least, at the contract level.

@@

\begin{lstlisting}
type either = Left of int | Right of string
\end{lstlisting}

\begin{lstlisting}
{ tag : <Left>, val : Num} | { tag : <Right>, val : Str}
\end{lstlisting}

@@

\begin{lstlisting}
switch x.tag {
  Left => x.val + 1,
  Right => strings.fromNum x.val
}
\end{lstlisting}

Flow typing, implemented in Racket\cite{FlowTypingRacket1, FlowTypingRacket2},
Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
TypeScript\cite{FlowTypingTypeScript}, is capable of refining appropriately the
type information in each branch. Thus equipped with untagged unions and a
typechecker smart enough, classic tagged unions can be simulated, but with the
additional benefit that the language is not imposing any representation.
Untagged unions empower the programmer to represent and handle a wider variety
of union idioms appearing in the wild.  Thus, adding union types to Nickel
sounds appealing.
